---
title: "Homework 1"
author: "Dmitry Panteleev"
date: "2024-04-03"
output:
  word_document: default
  pdf_document: default
  html_document: default
---
## Homework 1
1. Make EDA for data (5 points).
2. Build an ordination of objects using NMDS methods (descriptions, samples, etc.) (5 points).
3. Visualise the relationship between the resulting ordination and environmental parameters with functions envfit() and ordisufr()(5 points).
4. Draw conclusions about the most important factors (5 points).

*Data sources**
Trees on Barro Colorado (data from Condit et al. (2002), 'BCI' data, 'vegan' package).
Backtick: `r ceiling(pi)`
```{r}
library(tidyr)
library(dplyr)
library(broom)

library(tidyverse)

library(ggmap)
library(ggvegan)
library(vegan)
```

Download the data of "Trees on Barro Colorado"

```{r}
data(BCI.env)
data(BCI)
```


EDA

```{r}

name <- -1
count <- -1

for (i in colnames(BCI))
{
  sum_col <- sum(BCI[i])
  name <- c(name, i)
  count <- c(count, sum_col)
  #sums <- sum(BCI)
}

name <- name[2:length(name)]
count <- count[2:length(count)]
df_count <- data.frame(name,count)

ggplot(data=df_count, aes(x=name, y=count)) +
  geom_col()

sort(table(as.character(df_count$count)))

df_remove <- df_count$name[df_count$count <= 3]

df_new <- subset(BCI, select = ! (names(BCI) %in% df_remove)) 

df_new$other <- rowSums(BCI[,df_remove])

```


```{r}

colnames(BCI.env)
env_BCI <- BCI.env[,c("Age.cat","Habitat","Stream")]
print(is.factor(env_BCI$Habitat))

print(levels(env_BCI$Habitat))
print(levels(env_BCI$Age.cat))
print(levels(env_BCI$Stream))
```


Build an ordination of objects using NMDS methods

```{r}

ord_BCI <- metaMDS(df_new, dist = "bray")

ord_BCI_auto <- metaMDS(df_new, dist = "bray", autotransform = F)

pal_col <- c("red","orange", "yellow", "green", "steelblue")
pal_sh <- c(1, 2)


ordiplot(ord_BCI_auto, type = "point")
points(ord_BCI_auto, col = pal_col[env_BCI$Habitat], pch = pal_sh[env_BCI$Stream])

ordiplot(ord_BCI_auto, type = "n")
points(ord_BCI_auto, col = pal_col[env_BCI$Habitat], pch = pal_sh[env_BCI$Stream])
```
We have to use ord_BCI_auto because without autotransformation the stress is more than 0.2

```{r}

ordiplot(ord_BCI_auto, type = "n")
points(ord_BCI_auto, col = pal_col[env_BCI$Habitat], pch = pal_sh[env_BCI$Stream])
# Legend (example of relative and absolute positioning)
legend("topleft", bty = "n",
       title = "Habitat: ",
       legend = levels(env_BCI$Habitat), col = pal_col, pch = 15)
legend("topright", xjust = 1, yjust = 1,
       title = "Stream: ",
       legend = levels(env_BCI$Stream), col = "black", pch = pal_sh)
```

```{r}
op <- par(mar = c(3, 3, 0.1, 0.1), mgp = c(2, 1, 0))
ordiplot(ord_BCI_auto, type = "n")
points(ord_BCI_auto, col = pal_col[env_BCI$Habitat], pch = pal_sh[env_BCI$Stream])
legend("topleft", bty = "n",
       title = "Habitat: ",
       legend = levels(env_BCI$Habitat), col = pal_col, pch = 15)
legend("topright", xjust = 1, yjust = 1,
       title = "Stream: ",
       legend = levels(env_BCI$Stream), col = "black", pch = pal_sh)
text(ord_BCI_auto, display = "species", cex = 0.2, col = "grey20")
par(op)
```
## Envfit 


```{r}
ef <- envfit(ord_BCI_auto, env_BCI)

scores(ef, "vectors")

ef$vectors
ef$factors
colnames(env_BCI)

ordiplot(ord_BCI_auto, type = "n")
points(ord_BCI_auto, col = pal_col[env_BCI$Habitat], pch = pal_sh[env_BCI$Stream])
plot(ef, cex = 0.3)
```
I didn't get any vectors and I decided to change my variables from character to numeric
```{r}
levels(env_BCI$Habitat)

df <- env_BCI

df %>% mutate_if(is.factor, as.character) -> df
class(df$Habitat)
df['Habitat'][df['Habitat'] == 'Young'] <- 0
df['Habitat'][df['Habitat'] == 'Swamp'] <- 1
df['Habitat'][df['Habitat'] == 'OldSlope'] <- 2
df['Habitat'][df['Habitat'] == 'OldLow'] <- 3
df['Habitat'][df['Habitat'] == 'OldHigh'] <- 4

df['Stream'][df['Stream'] == 'No'] <- 0
df['Stream'][df['Stream'] == 'Yes'] <- 1

df['Age.cat'][df['Age.cat'] == 'c2'] <- 0
df['Age.cat'][df['Age.cat'] == 'c3'] <- 1

df$Habitat_f <- as.factor(df$Habitat)
levels(df$Habitat_f)

df$Stream_f <- as.factor(df$Stream)
levels(df$Stream_f)

df$Age.cat_f <- as.factor(df$Age.cat)
levels(df$Age.cat_f)


ef <- envfit(ord_BCI_auto, df[,c("Habitat_f", "Stream_f")])

scores(ef, "vectors")

ef$vectors
ef$factors
colnames(env_BCI)

ordiplot(ord_BCI_auto, type = "n")
points(ord_BCI_auto, col = pal_col[env_BCI$Habitat], pch = pal_sh[env_BCI$Stream])
plot(ef, cex = 0.3)


```
And here I didn't get vectors. I understood a problem, i need to use continuous variables:

```{r}

ef <- envfit(ord_BCI_auto, BCI.env[,c("EnvHet","UTM.NS","UTM.EW")])

scores(ef, "vectors")
levels(BCI.env$EnvHet)

ef$vectors
ef$factors

ordiplot(ord_BCI_auto, type = "n")
points(ord_BCI_auto, col = pal_col[env_BCI$Habitat], pch = pal_sh[env_BCI$Stream])
plot(ef, cex = 0.5)
```


Ordisurf:

```{r}
class(df$Habitat)
df$Habitat <- as.numeric(df$Habitat)
df$Age.cat <- as.numeric(df$Age.cat)
df$Stream <- as.numeric(df$Stream)

par(mfrow = c(1, 2))
os_EnvHet <- ordisurf(ord_BCI_auto, BCI.env$EnvHet, method = "REML")
os_UTM.NS <- ordisurf(ord_BCI_auto, BCI.env$UTM.NS, method = "REML")
os_UTM.EW <- ordisurf(ord_BCI_auto, BCI.env$UTM.EW, method = "REML")
os_EnvHet <- ordisurf(ord_BCI_auto, df$Habitat, method = "REML")
os_UTM.NS <- ordisurf(ord_BCI_auto, df$Stream, method = "REML")
os_UTM.EW <- ordisurf(ord_BCI_auto, df$Age.cat, method = "REML")

par(mfrow = c(1, 1))


```
# The result of these vectors:
1) EnvHet and UTM.EW have a strong positive correlation
2) Two factors (EnvHet, UTM.EW) have a moderate negative correlation with UTM.NS

The most strong environmental factor is UTM.EW 

# The result of surface plots 

We can see the complex non-linear impact of different factors on the amount of trees except UTM.EW, at this plot we can see approximate linear relationship
